Modelling multi-scale state-switching functional data with hidden Markov models

### 2021-05-18 Report/Sections/intro.tex
Updates from Overleaf

FDA was originally developed to process curves assumed to be independent replicates (i.e., there is no between-curve dependence), and within-curve fine-scale structure is not usually incorporated in FDA models. However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
%, as is the case for our case study data. Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time. 
For example, dive profiles of marine animals show distinct dive ``types" that cluster together in time \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}. Fine-scale periodicity within a larger process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.

@@

Some FDA models account for between-curve dependence occurring when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain types of dependence in time. For example, previous studies have used multilevel models with random effects to model variation between and within individuals in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}. More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}. However, these multilevel models are not appropriate in many biologging applications since they do not account for different curve types. Further, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence is not appropriate in most biologging applications. 
In addition to multi-level models, FDA researchers have also used functional time series to model dependence in a sequence of curves. Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by well-defined hidden states.

@@

To accommodate both temporal dependence and changes in curve type, we turn to the field of
%methodology for the analysis of animal movement 
%statistical ecology
animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}. HMMs interpret animal movement data as arising from a sequence of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position. While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions. In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, 
%This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments. 
but this approach does not account for certain types of temporal correlation.

@@

%While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together. 
In this paper, we combine existing methods from statistical ecology literature in novel ways to account for complex, temporally dependent functional data. The resulting suite of methods makes up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.
%and thus extend the methods available for the analysis of functional data. 
We begin in Section 2 by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures. We also show how these methods can be combined to analyze increasingly complex data. In Section 3 we fit several candidate models to data from a killer whale (\textit{Orcinus orca}) from the threatened northern resident population off the coast of British Columbia, Canada. Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.

>>>>

FDA was originally developed to process curves assumed to be independent replicates (i.e., there is no between-curve dependence), and within-curve fine-scale structure is not usually incorporated in FDA models. However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
%, as is the case for our case study data. Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time. 
For example, dive profiles of marine animals show a discrete number of distinct dive ``types" that aggregate together in time \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}. Fine-scale periodicity within a larger state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.

@@

Some FDA models account for between-curve dependence occurring when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain kinds of dependence in time. For example, previous studies have used multilevel models with random effects to model variation between and within individuals in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}. More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}. %However, these multilevel models are not appropriate in many biologging applications since they rely on  do not account for a discrete number curve types. 
However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time, which is not appropriate in most biologging applications where behaviours often change suddenly between a discrete number of types.
In addition to multi-level models, FDA researchers have also used functional time series to model dependence in a sequence of curves. Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by well-defined hidden states.

@@

We aim to identify and describe distinct curve types within sequential functional data while simultaneously modelling the complicated temporal dependence both between and within those curves. To accommodate both temporal dependence and sudden changes in behaviour, we turn to the field of
%methodology for the analysis of animal movement 
%statistical ecology
animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}. HMMs interpret animal movement data as arising from a Markov chain of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position. While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions. In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, 
%This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments. 
but this approach does not account for certain types of temporal correlation.

@@

%While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together. 
In this paper, we combine existing methods from statistical ecology literature in novel ways to classify and describe state-switching behaviours in functional data while accounting for complex temporal dependence. The resulting suite of methods makes up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.
%and thus extend the methods available for the analysis of functional data. 
We begin in Section 2 by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures. We also show how these methods can be combined to analyze increasingly complex data. In Section 3 we fit several candidate models to data from a killer whale (\textit{Orcinus orca}) from the threatened northern resident population off the coast of British Columbia, Canada. Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.

### 2021-05-18 Report/Sections/results.tex
Updates from Overleaf

We do not use information criteria to select the number of dive types $N$ since these metrics tend to overestimate the number of behavioural states in biological processes \citep{Pohle:2017}. We instead plot the duration of each dive versus the duration of the dive preceding it ($y_t$ versus $y_{t-1}$ for $t \geq 2$). This type of visualization is known as a lag plot. If the emission distributions of the hidden states are well-separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}. This is unfortunately not the case for our killer whale data, as there is one cluster of data centred at approximately $y_t = y_{t-1} = 30$ seconds. However, longer dives appear to be characterized by bouts of less ``wiggly" behaviour in the acceleration data compared to shorter dives, so we choose $N = 2$ to differentiate these dive types. The absence of a more principled method to select $N$ highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
%
Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Figure 1 of the supplementary material), and visual inspection shows no obvious complicated dependence. Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor a moving-window transformation is called for.
%
Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows a Gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
%
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
%
This is consistent with previous studies, including that of \citet{Barajas:2017}. 
%

@@

In total, the parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \theta = \{\mu^{(1)},\sigma^{(1)},\mu^{(2)},\sigma^{(2)}\} \qquad \text{($Y$ emission parameters), and} \\
    %
    \theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{($\Z$ emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)},\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}.
\end{gather*}
%
Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $\Z_{t,\tilde t^*}$ conditioned on $\tilde X^*_{t,\tilde t^*} = i^*$. 
%
We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation. The likelihood of this model is easily calculated using the forward algorithm and can be maximized with respect to the parameters above (see the appendix for details). Figure \ref{fig:CarHHMM-DFT} shows the dependence structure of the full CarHHMM-DFT.

@@

The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which likely reflect resting, travelling, and to a lesser extent searching for prey.
Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives. No dive in this data set has a maximum depth deeper than 30 meters and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and significant numbers of fish have been observed to be caught near the surface (Fortune and Trites, unpublished data). Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.

@@

The estimated probability transition matrices and associated stationary distributions on the coarse scale are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
%
for the transitions between dives. The estimated probability transition matrices and stationary distributions on the fine scale are 
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for dive types 1 and 2.
%The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that 
In summary, about 79\% of dives are short dives of type 1, and
%The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that 
the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive. %The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that 
Further, this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 compared to only only 9\% of the time during a dive of type 2. Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.

>>>>

We do not use information criteria to select the number of dive types $N$ since these metrics tend to overestimate the number of behavioural states in biological processes \citep{Pohle:2017}. We instead plot the duration of each dive versus the duration of the dive preceding it ($y_t$ versus $y_{t-1}$ for $t \geq 2$). This type of visualization is known as a lag plot. If the emission distributions of the hidden states are well-separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}. This is unfortunately not the case for our killer whale data, as there is one cluster of data centred at approximately $y_t = y_{t-1} = 30$ seconds. However, longer dives appear to be characterized by bouts of less ``wiggly" behaviour in the acceleration data compared to shorter dives, so we choose $N = 2$ to differentiate these dive types. The absence of a more principled method to select $N$ highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
%
Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Figure 1 of the supplementary material), and visual inspection shows no obvious complicated dependence. Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor a moving-window transformation is called for.
%
Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows a Gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:

\[\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.\]
%
This is consistent with previous studies, including that of \citet{Barajas:2017}. 
%

@@

In total, the parameters to estimate are

\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \theta = \{\mu^{(1)},\sigma^{(1)},\mu^{(2)},\sigma^{(2)}\} \qquad \text{($Y$ emission parameters), and} \\
    %
    \theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{($\Z$ emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)},\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}.
\end{gather*}
%
Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $\Z_{t,\tilde t^*}$ conditioned on $\tilde X^*_{t,\tilde t^*} = i^*$. 
%
We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation. The likelihood of this model is easily calculated using the forward algorithm and can be maximized with respect to the parameters above (see the appendix for details). Figure \ref{fig:CarHHMM-DFT} shows the dependence structure of the full CarHHMM-DFT.

@@

The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which likely reflect resting, travelling, and to a lesser extent searching for prey.
Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives. No dive in this data set has a maximum depth deeper than 30 meters and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and significant numbers of fish have been observed to be caught near the surface \citep{Fortune:2020}. Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.

@@

The estimated probability transition matrices and associated stationary distributions on the coarse scale are

\[\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }\]

\[\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}\]
%
for the transitions between dives. The estimated probability transition matrices and stationary distributions on the fine scale are

\[\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},\]

\[\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}\]
%
for dive types 1 and 2.
%The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that 
In summary, about 79\% of dives are short dives of type 1, and
%The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that 
the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive. %The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that 
Further, this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 compared to only only 9\% of the time during a dive of type 2. Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.
